Tracing the cultural evolution of meaning at the semantics-pragmatics interface

### 2016-11-20 paper/paper-v2.tex
paper rewrite section 2 first pass finished

The replicator equation gives us the means to make the ensuing dynamics precise, without
necessarily committing to a biological or cultural interpretation. The proportion of types in a
given population is codified in a vector $\vec{x}$, where $x_i$ is type $i$'s proportion. The
fitness of type $i$ is type $i$'s average expected communicative success, or \emph{expected
  utility} (EU), given the current frequencies of types in the current population:
\begin{align*}
  f_i = \sum_j x_j \text{EU}(t_i,t_j)\,.
\end{align*}
The expected utility $\text{EU}(t_i,t_j)$ for type $i$ when communicating with type $j$ is the
average success of $i$ when talking or listening to $j$. Assuming that agents are speakers half
of the time this yields:
\begin{align*}
  \text{EU}(t_i,t_j) = \nicefrac{1}{2} \, U_S(t_i,t_j) + \nicefrac{1}{2} \, U_H(t_i,t_j)\,,
\end{align*}
where $\text{EU}_S(t_i,t_j)$ and $\text{EU}_H(t_i,t_j)$ are the expected utilities for $i$ as a
speaker and as a hearer when communicating with $j$, defined as follows where $n_i$ and $n_j$
are type $i$'s and $j$'s pragmatic reasoning types and $L_i$ and $L_j$ are their lexica:
\begin{flalign*}
  & \text{EU}_S(t_i,t_j)  = \sum_s P(s)\sum_m S_{n_i}(m \mid s;L_i) \sum_{s'} R_{n_j}(s' \mid m;L_j)
  \delta(s,s') \\
 & \text{EU}_H(t_i,t_j)  = \text{EU}_S(t_j,t_i)
\end{flalign*}
As usual, $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise. % The average fitness of the
% population is given by $\Phi$, $\Phi = \sum_i x_i f_i$, serving as a normalizing constant for
% the discrete replicator equation: $x'_i = \frac{x_i f_i}{\Phi}$.

\bigskip

In their series of papers on language evolution, Nowak and colleagues did not only consider
expressivity but also recognized the central role of the fidelity by which language is
transmitted. Among others, linguistic production can be prone to errors, states or messages may
be percieved incorrectly, and multiple languages may be compatible with the data learners are
exposed to. These sources of uncertainty introduce variation in their transmission from one
generation to the next which can be likened to mutation to the effect that a type's offspring
may adopt a different type than that of its parent as a result of transmission
perturbations. Importantly, this variation should be relative to the learnability of a type
under these perturbations (instead of, e.g., being a constant that is equal across all types as
in \citealt{nowak+etal:2002}). For this purpose, we turn to a different strand of research in
cultural evolution: iterated learning.

%errors were considered in N+K:1999, mutation in N+etal:2002

@@

Iterated learning is a process in which the behavior of one individual serves as learning input
for another. This learner, upon acquisition of this behavior, then goes on to produce behavior
that serves as input for a new learner. This process can be thought of as a progression through
chains of parents and children; the parent produces linguistic data from which the child infers
a language. The latter, now a parent, goes on to produce data for a new generation of
learners. Following \citet{griffiths+kalish:2007} we model iterated learning as a repeated
process of Bayesian inference in which learners combine the likelihood of a type producing the
received learning input with prior inductive biases.

Due to the pressure towards learnability it exherts, iterated learning generally leads to
simpler and more regular languages (see \citealt{kirby+etal:2014} and
\citealt{tamariz+kirby:2016} for recent surveys). Importantly, experimental and mathematical
results have been taken to suggest that the outcome of this process reflects learners'
inductive biases. In a Bayesian setting these biases can be codified in a prior
$P \in \Delta(T)$, which can be thought of as the amount of data a learner would require in
order to adopt a language (cf. \citealt[450]{griffiths+kalish:2007}). Or, in our case, a
combination of a lexicon and its use. The extent of the prior's influence has been shown to
heavily depend on the learning strategy assumed to underly the inference process. On the one
hand, early simulation results suggested that weak biases could be magnified by exposing
learners to only small data samples (e.g. in \citealt{brighton:2002}). On the other,
\citeposs{griffiths+kalish:2007} mathematical characterization showed that iterated learning
converged to the prior, i.e., the resulting distribution over languages corresponds to the
learners' prior distribution and is not influenced by the amount of input given to them. This
difference in predictions can be traced back to differences in the selection of hypotheses from
the posterior. Griffith \& Kalish's convergence to the prior holds for learners that sample
from the posterior. More deterministic strategies such as the adoption of the type with the
highest posterior probability, so-called {\it maximum a posterior estimation} (MAP), increase
the influence of both the prior and the data \citep{griffiths+kalish:2007,kirby+etal:2007}. In
the following, we parametrize the posterior by $l \geq 1$. In doing so, we obtain a range of
learning strategies between posterior sampling and MAP. When $l = 1$ learners sample from the
posterior and the learners propensity to maximize the posterior grows as $l$ increases.

The data learners are exposed to is described by a set $D$ containing sequences of state-message pairs, e.g., $\tuple{\tuple{s_i,m_v},...\tuple{s_j,m_w}}$. These are sequences of language use witnessed by learners, the length of which we denote by $k$. %Put differently, each datum $d \in D$ contains $k$ members of the set $\{\tuple{s_i,m_v} | s_i \in S, m_v \in M\}$.

%To summarize, the parametrized posterior, $P(d|t)^l$, is obtained from combining the prior $P(t)$ and the likelihood $P(d|t)$. Since types are combinations of a lexicon and its use, the latter can be computed straightforwardly from a type's production behavior. 
We combine the above with the replicator dynamics by codifying iterated learning in a transition matrix $Q$, where $Q_{ij}$ indicates the probability that a child of a parent of type $i$ adopts type $j$. This quantity is proportional to the probability of $i$ producing the learning data and that of inferring $j$ given the data:

\[
 Q_{ij} \propto \sum_{d \in D} P(d|t_i) F(t_j,d), \; \text{where } F(t_j,d) \propto P(t_j|d)^l \text{ and } P(t_j|d) \propto P(t_j) P(d|t_j).
\]

\subsection{Model summary}
Drawing from past research, we argued that expressivity and learnability are central to the cultural evolution of language. We propose these components to be modelled respectively as communicative efficiency-relative replication and (iterated) Bayesian learning. Taken together their interaction is described by the replicator-mutator dynamics \citep{hofbauer+sigmund:2003}:

\[ 
\hat{x}_i = \sum_j Q_{ji} \frac{x_jf_j}{\Phi}
\]

The units that the dynamics operate on are a combination of a lexicon and a degree of pragmatic sophistication determining its use. We call this combination a type. A type's expressivity depends on its communicative efficiency within a population while its learnability depends on the fidelity by which it is inferred by new generations of learners. The learners' task is consequently to perform a joint inference over types of linguistic behavior and lexical meaning. With this model at hand we turn to the analysis of the lack of lexicalization of productive pragmatic inferences in a case study on scalar implicatures.

>>>>

The replicator equation gives us the means to make the ensuing dynamics precise, without
necessarily committing to a biological or cultural interpretation. The proportion of types in a
given population is codified in a vector $\vec{x}$, where $x_i$ is type $i$'s proportion. The
fitness of type $i$ is type $i$'s average expected communicative success, or \emph{expected
  utility} (EU), given the current frequencies of types in the current population:
\begin{align*}
  f_i = \sum_j x_j \text{EU}(t_i,t_j)\,.
\end{align*}
The expected utility $\text{EU}(t_i,t_j)$ for type $i$ when communicating with type $j$ is the
average success of $i$ when talking or listening to $j$. Assuming that agents are speakers half
of the time this yields:
\begin{align*}
  \text{EU}(t_i,t_j) = \nicefrac{1}{2} \, U_S(t_i,t_j) + \nicefrac{1}{2} \, U_H(t_i,t_j)\,,
\end{align*}
where $\text{EU}_S(t_i,t_j)$ and $\text{EU}_H(t_i,t_j)$ are the expected utilities for $i$ as a
speaker and as a hearer when communicating with $j$, defined as follows where $n_i$ and $n_j$
are type $i$'s and $j$'s pragmatic reasoning types and $L_i$ and $L_j$ are their lexica:
\begin{flalign*}
  & \text{EU}_S(t_i,t_j)  = \sum_s P(s)\sum_m S_{n_i}(m \mid s;L_i) \sum_{s'} R_{n_j}(s' \mid m;L_j)
  \delta(s,s') \\
 & \text{EU}_H(t_i,t_j)  = \text{EU}_S(t_j,t_i)
\end{flalign*}
As usual, $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.

@@

Languages are shaped not only by functionalist forces towards greater expressivity. Another
important factor is the fidelity by which language is transmitted. Among others, linguistic
production can be prone to errors, states or messages may be perceived incorrectly, and
multiple languages may be compatible with the data learners are exposed to. These sources of
uncertainty introduce variation in their transmission from one generation to the next. Biases
or systematic noise in the iterated transmission process can influence language evolution
substantially.

In biological evolution, where types are expressed genetically, transmission infidelity comes
into the picture through infrequent and mostly random genetic mutations. But an agent's lexicon
and pragmatic reasoning behavior is not inherited genetically. They need to be learned from
observation. Concretely, when agents of type $j$ want to adopt or imitate the linguistic
behavior of type $i$, they observe the linguistic behavior of type $i$ and need to infer what
their type is from that. Iterated learning is a process in which languages are learned repeatedly from the observation
of linguistic behavior of agents who have acquired the language from observation and inference
in the same way before. In the simplest case, there is a single teacher and a single
learner. After sufficient training, each learner becomes a teacher and produces behavior that
serves as input for a new learner. Due to the pressure towards learnability it exerts, iterated
learning generally leads to simpler and more regular languages (see \citealt{kirby+etal:2014}
and \citealt{tamariz+kirby:2016} for recent surveys).

Following \citet{griffiths+kalish:2007} we model language acquisition as a process of Bayesian
inference in which learners combine the likelihood of a type producing the received learning
input with prior inductive biases. Experimental and mathematical results on iterated learning
suggest that the outcome of this process reflects learners' inductive biases
\citep[e.g.][]{kirby+etal:2014}. In a Bayesian setting these biases can be codified in a prior
$P \in \Delta(T)$, which reflect the amount of data a learner requires to faithfully acquire
the language of the teacher \citep[c.f.][450]{griffiths+kalish:2007}. The extent of the prior's
influence has been shown to heavily depend on the learning strategy assumed to underly the
inference process. On the one hand, early simulation results suggested that weak biases could
be magnified by exposing learners to only small data samples (e.g. in
\citealt{brighton:2002}). On the other, \citeposs{griffiths+kalish:2007} mathematical
characterization showed that iterated learning converged to the prior, i.e., the resulting
distribution over languages corresponds to the learners' prior distribution and is not
influenced by the amount of input given to them. This difference in predictions can be traced
back to differences in the selection of hypotheses from the posterior. Griffith \& Kalish's
convergence to the prior holds for learners that sample from the posterior. More deterministic
strategies such as the adoption of the type with the highest posterior probability, so-called
{\it maximum a posterior estimation} (MAP), increase the influence of both the prior and the
data \citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we use a paremeter
$l\ge1$ to modulate between posterior sampling and the MAP strategy. When $l = 1$ learners
sample from the posterior and the learners propensity to maximize the posterior grows as $l$
increases.

Let $D$ be the set of possible data that learners may be exposed to. This set $D$ contains all
sequences of length $k$ of state-message pairs, e.g.,
$\tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}}$. The number $k$ of observations is another model
parameters. As $k$ increases, learners have more data to base their inference on and so tend to
recover the true types that generated a given sequence with higher probability. The mutation
matrix $Q$ of the replicator mutator dynamics in (\ref{eq:RMD_discrete}) can then be defined as
follows: $Q_{ji}$ is the probability that a learner acquires type $i$ when learning from an
agent of type $j$. The learner observes one length-$k$ sequence $d$ of state-message pairs, but
the probability $P(d \mid t_j)$ with which sequence $d = \tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}}$ is observed depends on type $j$'s
behavior:
\begin{align*}
  P(d = \tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}} \mid t_j) = \prod_{i = 1}^k S_{n_j}(m_i
  \mid t_i; L_{j})\,,
\end{align*}
where, as before, $n_j$ is $j$'s pragmatic reasoning type and $L_j$ is $j$'s lexicon. For a
given observation $d$, the probability of acquiring type $i$ is $F(t_i \mid d)$, so that:
\mf{check if we need ``proportional to'' here or whether ``equals to'' is okay}
\begin{flalign*}
  Q_{ji} \propto \sum_{d \in D} P(d \mid t_j) F(t_i \mid d)\,.
\end{flalign*}
The acquisition probability $F(t_i \mid
d)$ given datum $d$ is obtained by probability matching $l = 1$ or a tendency towards choosing
the most likely type $l > 1$ from the posterior distribution $P(\cdot \mid d)$ over types given
the data, which is calculated by Bayes' rule:
\begin{flalign*}
  & F(t_i \mid d) \propto P(t_i \mid d)^l \; \text{ and }\\
  & P(t_i \mid d) \propto P(t_i) P(d \mid t_i)\,.
\end{flalign*}

\subsection{Model summary}

Expressivity and learnability are central to the cultural evolution of language. These
components can be modelled, respectively, as replication based on a measure of fitness in terms
of communicative efficiency and iterated Bayesian learning. Their interaction is described by
the discrete time replicator mutator dynamics in (\ref{eq:RMD_discrete}), repeated here:
\begin{align*}
  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_k x_k f_k}\,.
\end{align*}
This equation defines the frequency $x'_i$ of type $i$ at the next time step, based on its
frequency $x_i$ before the update, its fitness $f_i$ and the probability that a learner infers
$i$ when observing the behavior of a type-$j$ agent. Fitness-based selection can be thought of
as biological (fitness as expected relative number of offspring) or cultural (fitness of
likelihood of being imitated or repeated). The types that the dynamic operates on are pairs
consisting of a lexicon and a pragmatic use pattern. A type's expressivity depends on its
communicative efficiency within a population while its learnability depends on the fidelity by
which it is inferred by new generations of learners. The learners' task is consequently to
perform a joint inference over types of linguistic behavior and lexical meaning.
